---
title: "Covid"
author: "Kevien Otieno"
date: "2023-10-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#loading packages
```{r}
pacman::p_load(dplyr, tidyr, tidyverse, janitor, ggplot2, readxl,VennDiagram, lubridate)
```

#Covid
```{r}
Covid = read.csv("Covid 19-studies.csv",
                   check.names = F,
                   na= "")
```

```{r}
#converting sex, Age and Study Status columns to factor
Covid$Sex = as.factor(Covid$Sex)
Covid$Age = as.factor(Covid$Age)
Covid$`Study Status` = as.factor(Covid$`Study Status`)
```

```{r}
# Create an empty vector to store the country names
country <- vector("character", length = nrow(Covid))

# Loop through each row
for (i in 1:nrow(Covid)) {
  # Split the "Locations" column for the current row
  row_splits <- unlist(str_split(Covid$Locations[i], ","))
  
  # Calculate the number of countries for the current row
  length_loc <- length(row_splits)
  
  # Store the country name for the current row
  country[i] <- row_splits[length_loc]
}
```

```{r}
#creating a countries dataframe
country_df= data.frame(country)
```

```{r}
#grouping The column countries by the countries
Country_study =country_df %>% 
  group_by(country) %>% 
  summarise(`No of Studies` = n())
#arranging the dataset in descending order using the No of studies
Country_study = Country_study[order(Country_study$`No of Studies`, decreasing = TRUE), ]
#removing Nas
Country_study = Country_study %>% 
  filter(!is.na(Country_study$country))
Country_study_top =head(Country_study, 10)
```

```{r}
ggplot(Country_study_top, aes(x=`No of Studies`, y=country,fill =country))+
  geom_bar(stat="identity", col= "black")+
  coord_cartesian(xlim = c(0,max(Country_study_top$`No of Studies`)))+
  labs(title = "Top 10 countries doing clinical studies")+
  theme_minimal()
```
From the studies conducted, United States leads on the No of studies being conducted and In Africa the leading country is only Egypt.

#To check if no of studies have been increasing over the years
```{r}
# Create an empty vector to store the years
year <- vector("character", length = nrow(Covid)) 

for (i in 1:nrow(Covid)) {
  # Split the "Locations" column for the current row
  row_splits1 <- unlist(str_split(Covid$`Start Date`[i], "-"))
  
  # Store the country name for the current row
  year[i] <- row_splits1[1]
}
```

```{r}
#converting the year vector to a dataframe
year1 =data.frame(year)

#merging the year dataframe with the Covid dataframe
Covid = cbind(Covid, year1)
```

```{r}
#converting year to numeric
Covid$year <- as.numeric(Covid$year)
```

```{r}
#grouping the years to the know the No of studies conducted each year
a= c(17,18)
year_studies = Covid %>% 
  group_by(year) %>% 
  summarise(`No of studies` = n()) %>% 
  filter(!is.na(year))
year_studies1=  year_studies[-a, ]
```

```{r,}
#a line plot to show No of studies over years
ggplot(year_studies1, aes(x = year, y = `No of studies`)) +
  geom_point()+
  #geom_smooth(method = "loess") +
  geom_smooth(aes(group=1))+
  labs(title = "A trend showing the increase in the No of studies over the years") +
  theme_minimal()
```

The graph is constant for some years and starts to rise from 2016 and peaks at 2020 this is due to the outbreak of COVID 19 causing a great catastrophic on the whole world then starts to decline this is after the pandemic.

#conditons
```{r}
#extracting Conditons and calculating the number of each
Conditions = Covid %>% 
  group_by(Conditions) %>% 
  summarise(`No of studies` = n()) 
#sorting the conditions by the No of studies
Conditions = Conditions[order(Conditions$`No of studies`, 
                              decreasing = TRUE), ]
#top 5 conditions are being studied
plot_conditions =head(Conditions, 5)
```

```{r}
#bar graph to show the leading condition
ggplot(plot_conditions, aes(x= `No of studies`, y = Conditions, fill= Conditions))+
  geom_bar(stat = "identity", col ="black")+
  coord_cartesian()+
  labs(Title= "Top five conditions being studied")+
  theme_minimal()
```
From graph, Most studies involve study on Covid 19 due to its catastrophic effect on the World.

#Interventions
```{r}
#grouping the interventions
Interventions = Covid %>% 
  group_by(Interventions) %>% 
  summarise(`No of studies` = n())
#sorting the interventions using the No of Studies
Interventions = Interventions[order(Interventions$`No of studies` , decreasing = TRUE), ]
```

```{r}
#getting the top 10 interventions
Interventions_plot = head(Interventions, 10)
```

```{r}
#Bar plot to clearly display the interventions mostly used
ggplot(Interventions_plot, aes(x=`No of studies`, y=Interventions))+
  geom_bar(stat = "identity",fill=c("red","blue","green","gray","black","yellow","purple","pink","navy","orange"))+
  coord_cartesian()+
  labs(title = "Most common interventions", x= "Interventions", y ="No of Studies")+
  theme_minimal()
```

#Major Sponsors for the different clinical trials
```{r}
#Grouping the sponsors
Sponsors = Covid %>% 
  group_by(Sponsor) %>% 
  summarise(`No of studies` =n())
#Sorting the sponsors according to the no of studies
Sponsors = Sponsors[order(Sponsors$`No of studies`, decreasing = TRUE), ]
```

```{r}
Sponsors_plot = head(Sponsors, 5)
```

#clear display of the lead sponsors
```{r}
ggplot(Sponsors_plot, aes(x= `No of studies`, y= Sponsor))+
  geom_bar(stat ="identity", fill= c("red","blue","green","gray","black"))+
  coord_cartesian()+
  labs(title ="Lead sponsors for the different clinical studies")+
  theme_minimal()
```


#colaborators
```{r}
#grouping the colaborators
colaborator= Covid %>% 
  group_by(Collaborators) %>% 
  summarise(`No of studies`= n())
#sorting the colaborators in descending order
colaborator= colaborator[order(colaborator$`No of studies`, decreasing = TRUE), ]
```

```{r}
#obtaining the top 5 colaborators
top_colaborators= head(colaborator, 5)
```
```{r}
#bar plot for a clear visual
ggplot(top_colaborators, aes(x=`No of studies`, y= Collaborators ))+
  geom_bar(stat = "identity",fill= c("red","blue","green","gray","black"))+
  coord_cartesian(xlim = c(0, max(top_colaborators$`No of studies`)))+
  labs(title= "Investigating the top colaborators")+
  theme_minimal()
```

The NA indicates that most studies lacked people collaborating with them

#Sex that is mostly involved in the study
```{r}
#Grouping people of different sex
Sex= Covid %>% 
  group_by(Sex) %>% 
  summarise(`No of studies`= n()) %>% 
  filter(!is.na(Sex))
```

```{r}
# Calculate percentages
Sex$percentage4 <- Sex$`No of studies` / sum(Sex$`No of studies`) * 100

# Create a pie chart
ggplot(Sex, aes(x = Sex, y = `No of studies`, fill = Sex)) +
  geom_bar(stat = "identity", width = 1) +
  geom_text(
    aes(label = paste0(round(percentage4, 1), "%")),
    position = position_stack(vjust = 0.5)
  )+
  labs(
    title = "Distribution of Studies by Sex",
    fill = "Sex"
  ) +
  theme(legend.position = "right") +
  scale_fill_manual(values = c("ALL" = "red", "FEMALE" = "blue", "MALE" = "green"))

```
From the pie chart most studies involved a combination of all genders.

#Investigating which Kind of age group are mostly involved in studies
```{r}
#grouping the age groups
age_group = Covid %>% 
  group_by(Age) %>% 
  summarise(`No of studies`= n())
percentage = ((age_group$`No of studies`)/sum(age_group$`No of studies`, na.rm = TRUE))*100
```

```{r}
# Create a pie chart
ggplot(age_group, aes(x = `No of studies`, y = Age, fill = Age)) +
  geom_bar(stat = "identity", width = 1) +
  geom_text(
    aes(label = paste0(round(percentage, 1), "%")),
    position = position_stack(vjust = 0.5)
  )+
  labs(
    title = "Distribution of Studies by Age groups",
    fill = "Age group"
  ) +
  theme(legend.position = "right")

```
From the graph most studies involved adults

#Investigating the study status
```{r}
#grouping the data according to Study Status
study_status= Covid %>% 
  group_by(`Study Status`) %>% 
  summarise(`No of Studies`= n())
#sorting No of studies in descending order
study_status= study_status[order(study_status$`No of Studies`, decreasing = TRUE), ]

#obtaining percentages for the different study status
percentage1= ((study_status$`No of Studies`)/sum(study_status$`No of Studies`))*100
```
```{r}
ggplot(study_status, aes(x=`No of Studies`, y= `Study Status`, fill =`Study Status`))+
  geom_bar(stat = "identity")+
  coord_cartesian()+
  geom_text(aes(label= paste0(round(percentage1, 1), "%"), vjust = 0.5, hjust = 0.1))+
  labs(title = "Investigating the study status")+
  theme_minimal()

```

We can see that most studies were completed

#Funder Type
```{r}
#grouping the Funder types
Funder.Type = Covid %>% 
  group_by(`Funder Type`) %>% 
  summarise(`No of Studies`=n())
#obtaining the percentages
percentage2= ((Funder.Type$`No of Studies`)/sum(Funder.Type$`No of Studies`))*100
```
```{r}
ggplot(Funder.Type, aes(x=`Funder Type`, y = `No of Studies`, fill = `Funder Type`))+
  geom_bar(stat = "identity")+
  geom_text(aes(label=(paste0(round(percentage2,1),"%")),vjust = 0.2, hjust= 0.5))+
  labs(title = "Funder for te different dtudies Conducted")+
  theme_minimal()

```

#study type
```{r}
Study_type= Covid %>% 
  group_by(`Study Type`) %>% 
  summarise(`No of Studies`= n())
```
```{r}
# Calculate percentages
percentage5 <- Study_type$`No of Studies` / sum(Study_type$`No of Studies`) * 100

# Create a pie chart
ggplot(Study_type, aes(x = "", y = `No of Studies`, fill = `Study Type`)) +
  geom_bar(stat = "identity", width = 1) +
  geom_text(
    aes(label = paste0(round(percentage5, 1), "%")),
    position = position_stack(vjust = 0.5)
  )+
  coord_polar("y") +
  labs(
    title = "Distribution of Studies by study type",
    fill = "`Study Type`"
  ) +
  theme_void() +
  theme(legend.position = "right")+
  scale_fill_manual(values= c("INTERVENTIONAL" = "blue", "OBSERVATIONAL" = "green","EXPANDED_ACCESS" = "red"))
```

From the graph most studies involved the use of a drug or Interventional.

#phases
```{r}
#grouping the data to different phases
Phases = Covid %>% 
  group_by(Phases) %>% 
  summarise(`No of Studies`= n()) %>% 
  filter(!is.na(Phases)) %>% 
  filter(Phases != "NA")
#obtaining the percentages
percentage3= ((Phases$`No of Studies`)/sum(Phases$`No of Studies`))*100
```

```{r}
ggplot(Phases, aes(x= `No of Studies`, y= Phases, fill= Phases))+
  geom_bar(stat = "identity")+
  coord_cartesian()+
  geom_text(aes(label=(paste0(round(percentage3,1),"%")),vjust = 0.2, hjust= 0.5))+
  labs(title = "Checking different studies in the study")+
  theme_minimal()
  

```
#Study results
```{r}
#grouping the study results
Study_results = Covid %>% 
  group_by(`Study Results`) %>% 
  summarise(`No of Studies`= n())
#obtaining the percentages
percentage4= ((Study_results$`No of Studies`)/sum(Study_results$`No of Studies`))*100
```
```{r}
ggplot(Study_results, aes(x=`Study Results`, y= `No of Studies`, fill =`Study Results`))+
  geom_bar(stat = "identity", )+
  geom_text(aes(label=(paste0(round(percentage4,1),"%")),vjust = 0, hjust= 0.5))+
  labs(title = "No of studies with results")+
  scale_fill_manual(values = c("NO"= "blue", "YES"= "green"))
```

From the graph we can clearly see that only a few of the studies yielded results.

